---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, echo = FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figure/README-"
)
library(dplyr)
```

[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Travis-CI Build Status](https://travis-ci.org/mdsumner/spbabel.svg?branch=master)](https://travis-ci.org/mdsumner/spbabel)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/mdsumner/spbabel?branch=master&svg=true)](https://ci.appveyor.com/project/mdsumner/spbabel)


# spbabel: a tidy view of Spatial

Spbabel provides simple tools to flip between Spatial and tidy forms of data. This package aims assist in the ongoing development of tools for spatial data in R. There is limited use for users directly, though see examples below. 

## Installation

Install the package from CRAN: 

```{r,eval=FALSE}
install.packages("spbabel")
```

The development version can be installed directly from github: 

```{r,eval=FALSE}
devtools::install_github("mdsumner/spbabel")
```

## Formal spatial data in R

Spatial data in the `sp` package have a formal definition  (extending class `Spatial`) that is modelled on shapefiles, and close at least in spirit to the [Simple Features definition](https://github.com/edzer/sfr). See [What is Spatial in R?](https://github.com/mdsumner/spbabel/wiki/What-is-Spatial-in-R) for more details. 

Turning these data into long-form tables has been likened to making fish soup, which has a nice nod to the universal translator [babelfish](https://en.wikipedia.org/wiki/List_of_races_and_species_in_The_Hitchhiker%27s_Guide_to_the_Galaxy#Babel_fish). Soup may be thick and creamy as in the "twice cooked" *bisque* or clear as in the "boiled" *bouillon* and this analogy can be applied in many ways. 

The `spbabel` package tries to help by providing a more systematic encoding into the long-form with consistent naming and lossless ways to re-compose the original (or somewhat modified) objects. 

The long-form version is similar to that implemented in: 

* sp's `as()` coercion for  `SpatialLinesDataFrame` to `SpatialPointsDataFrame`
* rasters's `geom()`
* ggplot2's `fortify()`
* gris' normalized tables
 
# How does spbabel work

After quite a lot of experimentation the long-form single table of all coordinates, with object, branch, island-status, and order provides the best middle-ground for transferring between different representations of Spatial data. Tables are always based on the "tibble" since it's a much better data frame. 

The `sptable` function creates the table of coordinates with identifiers for object and branch, which is understood by `sptable<-` to "fortify" and `sp` for the reverse. 

The long-form table may seem like soup, but it's not meant to be seen for normal use. It's very easy to dump this to databases, or to ask spatial databases for this form.  There are other more normalized multi-table approaches as well - this is just a powerful lowest common denominator. 

We can tidy this up by encoding the geometry data into a geometry-column, into nested data frames,  or by normalizing to tables that store only one kind of data, or with recursive data structures such as lists of matrices. Each of these has strengths and weaknesses. Ultimately I want this to evolve into a fully-fledged set of tools for representing spatial/topological data in R, but still by leveraging existing code whereever possible. 


# sptable: a round-trip-able extension to fortify

The `sptable` function decomposes a Spatial object to a single table structured as a row for every coordinate in all the sub-geometries, including duplicated coordinates that close polygonal rings, close lines and shared vertices between objects. 

The `sp` function re-composes a Spatial object from a table, it auto-detects the topology by the matching column names: 

* SpatialPolygons: object_, branch_, island_, order_
* SpatialLines: object_, branch_, order_
* SpatialPoints: object_
* SpatialMultiPoints: object_, branch_

The `sp` function could include overrides to avoid these tests but it's *so easy* to modify a table to have the matches for the required topology it hardly seems worth it. 

Use the `sptable<-` replacement method to modify the underlying geometric attributes (here `x` and `y` is assumed no matter what coordinate system). 

```{r}
library(maptools)
data(wrld_simpl)
library(spbabel)
(oz <- subset(wrld_simpl, NAME == "Australia"))

## long-form encoding of objects
oztab <- sptable(oz)

##  make a copy to modify
woz <- oz
## modify the geometry on this object without separating the vertices from the objects
sptable(woz) <- sptable(woz) %>% mutate(x_ = x_ - 5)

# plot to compare 
plot(oz, col = "grey")
plot(woz, add = TRUE)
```

We can also restructure objects, by mutating the value of object to be the same as "branch" we get individual objects from each.


```{r}

pp <- sptable(wrld_simpl %>% subset(NAME == "Japan" | grepl("Korea", NAME)))
## explode (or "disaggregate"") objects to individual polygons
## here each branch becomes an object, and each object only has one branch
## (ignoring hole-vs-island status for now)
wone <- sp(pp %>% mutate(object_ = branch_), crs = attr(pp, "crs"))
op <- par(mfrow = c(2, 1), mar = rep(0, 4))
plot(sp(pp), col = grey(seq(0.3, 0.7, length = length(unique(pp$object)))))
plot(wone, col = sample(rainbow(nrow(wone), alpha = 0.6)), border = NA)
par(op)
```

The long-form table is also ready for ggplot2: 
```{r}
library(ggplot2)
ggplot(pp) + aes(x = x_, y = y_, fill = factor(object_), group = branch_) + geom_polygon()

ggplot(pp) + aes(x = x_, y = y_, fill = factor(branch_), group = branch_, col = object_) + geom_polygon()
```

Please note that that `geom_polygon` cannot handle islands with multiple holes, and it only can do one hole by pretending it is a closed path and hides the boundary so you don't see the sleight of hand.  Search for "geom_holygon" for a way forward within ggplot2. 

# Why do this? 

I want these things, and spbabel is the right compromise for where to start: 

* flexibility in the number and type/s of attribute stored as "coordinates", x, y, lon, lat, z, time, temperature, etc.
* shared vertices
* ability to store points, lines and areas together, sharing topology where appropriate
* provide a flexible basis for conversion between other formats.
* flexibility and ease of use
* integration with database engines and other systems
* integration with D3 via htmlwidgets, with shiny, and with gggeom ggvis or similar
* data-flow with dplyr piping as the engine behind a D3 web interface

The ability to use [Manifold System](http://www.georeference.org/doc/manifold.htm) seamlessly with R is a particular long-term goal, and this will be best done(TM) via dplyr "back-ending". 

