---
title: "Geometry tables - read on-demand"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geometry tables read on-demand}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Another vignette in this package sent us a file. 

```{r,eval=FALSE}
library(dplyr)
db <- src_sqlite("data-raw/maps_county.sqlite3", create = FALSE)
db
```

There are two tables in it, and luckily we know what to do. 

```{r,eval=FALSE}
tbl(db, "geom")

tbl(db, "metadata")
```


Let's choose "wyoming", with "big horn" and "johnson". 

```{r,eval=FALSE}
meta <- tbl(db, "metadata")

meta %>% filter(state == "wyoming", county %in% c("big horn", "johnson")) %>% 
  select(county, group) %>% inner_join(tbl(db, "geom"))
```

Now we have something we can work with. 

## Read on-demand

```{r,eval=FALSE}
ggplot(
  meta %>% filter(state == "wyoming", county %in% c("big horn", "johnson")) %>% 
  select(county, group) %>% inner_join(tbl(db, "geom")) %>% 
    collect()
  
) + aes(x = long, y = lat, group = subregion, fill = subregion) + geom_polygon()
```

I'm interested now in the entire state. 

```{r,eval=FALSE}
ggplot(
  meta %>% filter(state == "wyoming") %>% 
  select(county, group) %>% inner_join(tbl(db, "geom")) %>% 
    collect() %>% group_by(subregion) %>% mutate(labx = mean(long), laby = mean(lat))
  
) + aes(x = long, y = lat, group = subregion, fill = subregion, label = subregion) + geom_polygon() + geom_path() + geom_text(aes(x = labx, y = laby)) +  
  scale_fill_grey(start = 0.3, guide = FALSE)
```


So, we have succesfully plotted our subset of polygons from a generic database containing two tables. 

Can we go the full distance and create a Spatial object? 

```{r,eval=FALSE}
library(spbabel)
  tbl(db, "metadata") %>% filter(state == "wyoming") %>% 
  select(county, group) %>% inner_join(tbl(db, "geom")) %>% 
    collect() %>% transmute(x_ = long, y_ = lat, branch_ = group, object_ = group, order_ = row_number(), island_ = TRUE, county = subregion, state = region) %>% sp(crs = "+proj=longlat +ellps=WGS84")

```


Obviously there's a lot of wasted coordinates here, so hopefully our friend figures out a more efficient storage in the db . . .

Show how the two tables may be combined, with nesting. 


## Simple features from normalized tables (mostly topology)


